Adaptation invariant concentration discrimination in an insect olfactory system

Version 1

Abstract

Abstract

Neural responses evoked by a stimulus reduce upon repetition. While this adaptation allows the sensory system to attend to novel cues, does information about the recurring stimulus particularly its intensity get compromised? We explored this issue in the locust olfactory system. We found that locusts’ innate behavioral response to odorants varied with repetition and stimulus intensity. Counter-intuitively, the stimulus-intensity dependent differences became significant only after adaptation had set in. Adaptation also altered responses of individual neurons in the antennal lobe (neural network downstream to insect antenna). These response variations to repetitions of the same stimulus were unpredictable and inconsistent across intensities. Although both adaptation and intensity decrements resulted in an overall reduction in spiking activities across neurons, these changes could be disentangled and information about stimulus intensity robustly maintained by ensemble neural responses. In sum, these results show how information about odor intensity can be preserved in an adaptation-invariant manner.

Introduction

The ability to adapt to sensory cues is key to the survival of many living organisms1–8. The goal of this sensory adaptation is to condense neural responses to a persisting or recurring stimulus in order to focus on novel cues. Such condensed neural responses are thought to increase system efficiency by reducing the metabolic requirements of information encoding9–12. However, is sensory information lost when neural responses are condensed during adaptation? While this computational task appears relatively elementary, any solution should satisfy two important requirements. First, attenuation of stimulus-evoked responses due to adaptation should not alter the identity of the stimulus. Second, information regarding the intensity of the stimulus may have to be preserved when odor intensity conveys information of behavioral importance. The latter requirement is paramount when sensory stimuli have different behavioral significance at varying intensities. For example, some odorants are attractive at lower intensities but switch to being aversive beyond certain threshold values13–15. Alternately, sensitivity to intensity gradients has to be maintained for sustained periods to allow for navigation towards or away from the stimulus source16–18. Such biological contexts necessitate the preservation of odor intensity information across repeated encounters with the odorant. However, the mechanism by which information about stimulus intensity is robustly maintained is not fully understood.

Sensory adaptation occurs over both long- and short-time scales12,19–21. Long-term adaptation requires gene transcription and protein translation that occurs on the time scale of hours to days6,22–24, whereas short-term adaptation occurs on a millisecond to minute time scale2,11,25. Short-term adaptation results from a wide range of mechanisms with prominent models in the olfactory system being vesicle depletion and facilitation of lateral inhibition25–32. While both vesicle depletion and lateral inhibition are likely to occur in the olfactory system, these complementary mechanisms will manifest in two different,but testable, outcomes. Vesicle depletion will impact the most prominent neurons with elevated odorevoked responses, whereas facilitation of lateral inhibition would progressively suppress activity in the relatively weak responders. Irrespective of which mechanisms underlie olfactory adaptation, the neural responses evoked by a stimulus will change over time and across repeated encounters. This raises the question, does adaptation then corrupt or alter information regarding the odorant?

The ambiguity that neural adaptation introduces has been documented in visual33 and auditory systems34,35. However, prior studies have shown that odor identity is robustly maintained by the spatiotemporally patterned neural responses and can allow precise recognition across repeated encounters with the odorant30,36–38. But is intensity information corrupted by the changes in neural activity due to sensory adaptation? Both reductions in stimulus intensity and sensory adaptation lead to a reduction in the overall spiking responses in most neural circuits. Do these neural response variations due to neural adaptation and intensity decrements be disentangled? We sought to examine this in the relatively simple insect olfactory system.

Our results show that odor intensity changes result in subtle variations in the overall combination of activated neurons. While sensory adaptation results in an overall reduction in the population neural responses, the activated ensemble is robustly maintained to allow simultaneous and precise decoding of odor identity and intensity. Using an innate behavioral assay, we show that physiological adaptation coincided with enhanced behavioral differentiation between odorant intensities. Thus, our work reveals a simple but elegant approach to achieve adaptation-invariant intensity coding and recognition in the insect olfactory system.

Results

Behavioral responses to an odorant vary with repetition and intensity

We began by examining how behavioral responses change with repetitions of an odorant. We used an innate behavioral assay where the presentation of certain odorants triggered an appetitive palp-opening response (POR) in starved locusts (Fig. 1a; Supp. Fig. 1). The probability of an odor eliciting a POR (P(POR)) was regarded as an indicator for stimulus perception (Fig. 1b). Note that the P(POR) was computed by averaging the behavioral responses across locusts for each trial or repetition number (Fig. 1c, d). Intriguingly, we found that the P(POR) systematically increased with repetitions of the same odorant and stabilized after about five stimulus presentations.

Behavioral response facilitation upon odor repetition.

(a)Schematic of the innate palp-opening response (POR) experimental paradigm. Opening of the locust’s maxillary palps within 15 s of the odor onset was considered a POR.

(b)Schematic of the block odor stimulation protocol. Each block consisted of ten trials and a 4s odor pulse was presented in each trial. A 15-minute no-stimulation period separated the blocks.

(c)Response matrices are shown summarizing individual locust PORs (rows) for blocks of ten trials (columns). White boxes indicate PORs in a specific trial, while black boxes indicate the absence of PORs in that trial. Locusts were sorted such that the least responsive locusts are shown at the top and the most responsive ones are near the bottom. PORs varied between locusts, across trials.

(d)The probability of PORs across locusts (P(POR)) is shown as a function of trial number for two odorants: iaa and bza at high intensities (1% v/v). P(POR)s as a function of trial number is shown for four odorants (oct, hex, iaa, bza) at two different intensities (0.1% v/v – low and 1% v/v – high). The trials were split into early (trials 1-5) and late (trials 6-10) trials. While the P(POR) of both odorant intensities increased over trials, the P(POR) values for high-intensity odor exposures were notably higher than P(PORs) for low-intensity odor presentations.

(f) Bar plots compare the average P(POR) for early and late trials for the low- and the high-intensity presentations of the same odor. P(POR) values observed at high and low intensities of an odorant were similar for early trials (p>0.05, one-tailed t-test). For three out of four odors tested the P(PORs) for high-intensity odor exposures were significantly higher (p<0.05, one-tailed t-test). This suggests that locusts can better perceptually distinguish between high and low odor intensities during later trials.

Previous results have shown that the POR is a function of both stimulus identity and intensity39–41. Therefore, we presented each locust with randomized blocks of the same odorant at either high (H, 1% v/v) or low (L, 0.1% v/v) intensities (Fig. 1e). Our results indicate that the P(POR) evoked by high and low stimulus intensities of an odorant were not significantly different in the earlier trials (Fig. 1f; p > 0.05 for hex, iaa, and bza; one-tailed t-test). However, the P(POR) became significantly different for the high and low-intensity exposures of the same odorant during the later trials (p ≤ 0.05 for hex, oct, and iaa; one-tailed t-test). In other words, high and low stimulus intensities were more behaviorally discriminable in the later trials. Hence, these results counterintuitively suggest that behavioral responses to an odorant at two different intensities were different after multiple encounters with the same stimulus.

Odor-evoked response adaptation leads to complex changes in individual neurons

How do neural responses change as a function of stimulus repetition? To understand this, we examined the odor-evoked responses of individual projection neurons (PNs) in the locust antennal lobe as different odorants at two different intensities were repeatedly presented (Fig. 2a). Odor-evoked responses in individual PNs varied subtly across trials (Fig. 2b). In general, most PN responses to the odorant were stronger in the first trial and reduced during later trials of the same stimulus. Notably, the changes in PN responses in two different blocks of trials, where the same odorant was presented, showed repeatable changes in odor-evoked spiking patterns and the overall response strength (Fig. 2c).

Ensemble neural activity systematically changes across repeated encounters with an odorant.

(a)Schematic of the olfactory stimulation protocol. Each block consisted of twenty-five trials with a four seconds odor pulse delivered in each trial. The inter-trial interval was 56 s. Two datasets were collected. Each dataset consisted of five randomized blocks of four odorants (dataset 1: hex (H), hex (L), oct (H), oct (L), hex (H)-repeat; dataset 2: iaa (H), iaa (L), bza (H), bza (L), iaa (H)-repeat). A fifteen-minute no-odor stimulation period separated blocks of trials.

(b)Raster plots of eight representative projection neurons (PNs) in the locust antennal lobe are shown. Spiking activities are shown for twenty-five trials (rows) with earlier trials shown at the top and later trials at the bottom. The shaded region indicates the four seconds odor stimulation period, and the identity of the stimulus is indicated in each plot.

(c)Raster plots are shown for four representative PNs during two blocks of trials. The same odorant was presented in both blocks. Note that spiking activity changes are repeatable across the block of trials.

(d) Schematic showing how vesicular depletion and lateral inhibition facilitation models would change spiking activity in individual neurons. Y-axis represents the change in response over trials ((1st trial response – 25th trial response)/ (1st trial response + 25th trial response)). Positive numbers indicate that the first trial had a stronger response, hence response reduction. Negative numbers indicate that the last trial had a stronger response, hence response facilitation. Along the x-axis, the response in the 1st trial is shown. Vesicular depletion should impact the strongly activated neurons more, whereas lateral inhibition facilitation should progressively suppress weak responders.

(e)Following the schematic in panel d, response changes observed in PNs are shown. Only PNs with significant odor-evoked responses are included in this plot. Responsiveness of PN was determined based on whether their peak odor-evoked response in the first trial exceeded the mean + 6.5 SDs of spontaneous activity preceding odor stimulation in that first trial. In total hex (L/H), n = 26, 39 PNs; oct (L/H), n = 35, 28 PNs; iaa (L/H), n = 39, 35 PNs; bza (L/H), n = 33, 40 PNs were included in the plot. Different colors and symbols are used to denote the identity of the odorant and its intensity (low-intensity trials – triangles and high-intensity trials – circles). The dotted line shows the regression fit revealing a linear relationship, if any, between the activation strength in the first trial (x-axis) and the response reduction observed (y-axis). As can be noted the R2 values are low indicating that both vesicle depletion and lateral inhibition facilitation models fail to capture the adaptation trends in our datasets.

(f)Changes in PN spiking activity over trials during high-intensity odor exposures (y-axis) are plotted against response changes observed for the same PN during low-intensity exposures of the same odorant. The dotted line indicates the regression fit revealing a linear relationship between the response changes observed during high and low-intensity exposures of the same odorant (hex, R2=0.034; oct, R2=0.032; iaa R2=0.067; bza R2=0.0016). The poor linear regression fit here indicates that reductions in neural response amplitude for one odor intensity do not model reductions in response amplitude for another odor intensity.

Next, we examined how individual PNs adapted during stimulus repetitions. Based on existing literature, two widely considered models of adaptation in the olfactory system are vesicular depletion and facilitation of lateral inhibition. While the vesicular depletion model predicts that the strongest firing neurons would adapt the most, in the lateral inhibition facilitation model the weakly activated neurons would progressively be suppressed more (Fig. 2d). To test these two opposing but complementary theories, we calculated the percentage change in the total spike counts observed during first and the 25th odor presentations (Fig. 2e, y-axis; positive values indicate weaker response in the later trial and hence response reduction, whereas negative values indicate facilitation of odor-evoked responses). We plotted these normalized response changes against the total spike counts observed in the very first trial (Fig. 2e, x-axis). The dotted lines represent the best linear model that fit the two datasets we recorded. If vesicular depletion was the predominant form of adaptation, the slope of the regression line would be positive. Whereas a negative slope would indicate that lateral inhibition was the predominant form of adaptation. However, as can be noted, across both datasets with two concentrations of the four odorants tested, neither model captured the adaptation dynamics in our datasets (R2 = 0.05, 0.01, respectively).

Are trial-to-trial changes in stimulus-evoked neural responses at two different concentrations of the same odorant correlated? If this is the case, then it would be reasonable to expect neurons that adapt more at the higher concentration to also change in a similar fashion at a lower concentration and vice versa. To assess this, we calculated and plotted the normalized change in spike counts for each PN at both high and low concentrations of the same odorant (Fig. 2f). Notably, the PNs with the greatest response change during repetitions of the higher-intensity odor exposures were not the ones that adapted heavily when encountering the same stimulus at a lower intensity (R2 = 0.034 (hex), 0.032 (oct), 0.067 (iaa), 0.0016 (bza)). In sum, these results indicate that change due to adaptation in individual neurons is not a simple function of its response strength. Further, how individual neurons adapted is a function of both odor identity and intensity.

Adaptation and intensity decrements both reduce ensemble neural response strength

What is the overall change in the spiking response across the neural ensemble? To examine this, we calculated the average spiking response across all PNs recorded and plotted the spike counts as a function of time for each trial (Fig. 3a). Expectedly, the overall response strength was stronger in the first trial and reduced when the same stimulus was repeatedly encountered. A similar reduction in overall response strengths was also observed when the odorant intensity was reduced (Fig. 3b)

Stimulus repetition and intensity decrements reduce spiking responses.

(a) Peristimulus time histograms (PSTHs) across all PNs (hex and oct, n = 80 PNs; iaa and bza, n = 81 PNs) for each odorant. PSTHs of trials 1 and 25 are shown.

(b) PSTHs across all PNs are compared between high and low-intensity odor exposures. Response during the first trial is shown.

(c) Summed spike counts (across PNs and four-second odor presentation) are calculated and shown as a function of the trial number. The dotted line indicates the spike count of the 25th trial of the high-intensity odor exposure.

Are the response reductions during stimulus repetition comparable to those observed with stimulus intensity decrements? To determine this, we plotted the total spike counts summed across all PNs over the entire stimulus presentation duration and plotted it as a function of the trial number (Fig. 3c). We compared the changes over trials with the total spike counts observed during the high (H) and low (L) intensities of the same odorant. Notably, the total PN ensemble spike counts decreased due to both stimulus intensity reduction and repetition of the stimulus. Do these results indicate that adaptation could potentially confound information regarding stimulus intensity?

Adaptation-invariant encoding of odor intensity

Could the information regarding stimulus intensity be robustly encoded in the population neural responses? As noted earlier, both spiking activities in individual neurons and the volume of spikes at the ensemble level vary across trials. If adaptation alters the overall spiking activity amplitude (i.e. vector length; Fig. 4a), while changes in stimulus intensity alter the combination of activated neurons (i.e. vector direction; Fig. 4b), then information about odor identity and intensity can be preserved in an adaptationinvariant fashion. To test whether the recorded ensemble of neurons in our datasets encoded information in such an adaption-invariant fashion, we first visualized the high-dimensional neural activities in each trial using a dimensionality reduction approach (see Methods). We plotted the ensemble responses in each 50 ms time bin during the odor presentation time window (4 s), and linked them based on the order of their occurrence, to generate trial-by-trial odor response trajectories (Fig. 4c, d). Note that each trial generated a single loop response trajectory after dimensionality reduction. Further, neural responses in different trials evoked closed-loop trajectories that systematically reduced in length with increasing repetition but maintained their direction indicating that the combination of neurons activated were mostly conserved. Different odorants evoked responses that were markedly different in the combination of activated neurons and therefore their response trajectories were oriented in different directions. Notably, different concentrations of the same odorant also generated response trajectories that subtly varied in direction. As a direct consequence, even though the overall spiking activities reduced over trials, the odor response trajectories reliably maintained their direction in all the trials. Hence, information about odor identity and intensity was robustly maintained across all trials.

Odor identity and intensity information is maintained across trials.

(a)A schematic showing how the ensemble neural activity might change between two odorants and across multiple trials or repetition. The combination of neural activated should differ between different odorants, and therefore the odor identity should be represented by population response vectors that differ in their direction. Repetitions should reduce response strength without altering the combination of neurons activated. If this were the case, the later repetitions of the same odorant would evoke a response that can be represented by vectors that maintain directions while progressively becoming lesser in length or magnitude.

(b)A schematic coding scheme for achieving adaptation invariant intensity coding. The combination of neurons activated changes markedly with odor identity and subtly with odor intensity. These responses become less intense without altering the combination of neurons activated. If this were the case, then the ensemble vector direction would change with both odor identity and intensity, and the vector direction robustly maintained even though the vector length continues to change with repetition.

(c)Trial-by-trial odor-evoked ensemble PN response trajectories after dimensionality-reduction are shown for hex and oct. Only high-concentration exposures of both odorants are included in this plot. For each odor, the temporal response trajectories for the 1st, 5th, 10th, 15th, 20th, and 25th trials are shown (color gradient from light to dark as trial number increases). Note that the odor-response trajectories become increasingly smaller during later trials, but the direction of the trajectories is robustly maintained.

(d)Trial-by-trial response trajectories for low concentrations exposures of the hex and oct are shown. Similar convention as panel c. (f)Similar plot as in panel c, but now comparing the odor-evoked response trajectories elicited by hex at high and low intensities.

(g)Similar plot as in panel d but comparing the response trajectories elicited by oct at high and low intensities.

To quantify these observations, we performed a correlation analysis using high-dimensional PN response vectors. If a similar combination of PNs (high-dimensional activity vectors) were activated in two different trials, then their correlation will be high. Correlations across different trials and concentrations of the same odorant are shown in Fig. 5a. Confirming results from the dimensionality reduction analysis (Fig. 4), we found odor-evoked responses across different trials of an odorant at a particular intensity were highly correlated (diagonal blocks on each panel). The similarity between odorevoked responses evoked by different concentrations of the same odorant was considerably less correlated (off-diagonal blocks). A hierarchical clustering analysis revealed that the odor-evoked responses nicely clustered based on odor identity and then by intensity (Fig. 5b)

High and low stimulus intensities activate distinct ensembles.

(a)Correlations between neural responses observed in different trials are shown. Each pixel/matrix-element represents a similarity between mean neural responses in one trial versus those in another trial. Diagonal blocks reveal the correlation between trials when the same odorant at a specific intensity was repeatedly presented.

(b)A dendrogram was generated using a correlation distance metric comparing trial-by-trial ensemble spiking activities evoked by two different stimuli at two different intensities (see Methods). Two major response clusters that correspond to stimulus identity and intensity were identified. The number at the leaf node represents the trial number.

In sum, our results reveal that a combinatorial code could encode information regarding odor identity and intensity in an adaptation-invariant fashion. Since the same information about a stimulus is represented with fewer spikes in the later trials, we conclude that adaptation refines the odor codes by making them more efficient.

Neural response suppression vs. behavioral output facilitation

Neural responses evoked by an odorant reduced with repetition (Fig. 3). However, as we noted earlier, the behavioral responses increased upon stimulus repetition (Fig. 1). This negative correlation is evident when the neural responses (total spiking activities evoked across neurons) was plotted against the behavioral POR response probability during a particular trial (Fig. 6). This negative correlation between neural suppression and behavioral enhancement was observed for all odorants and both concentrations examined. It is worth noting that the last trial of the high-intensity exposure evoked neural responses that were comparable in strength to the first trial of the low-intensity exposure of the same stimulus. Yet, the behavioral responses observed during these two trials were markedly different. Hence, our results also indicate that the probability of the POR response is not simply a function of the total number of spikes elicited.

Neural adaptation inversely correlates with behavioral facilitation.

(a) The probability of odor-evoked POR (P(POR)) for a given trial is plotted against the total spike counts elicited during the trial or repetition number. Therefore each point represents a single trial. Symbols with light and dark colors are used for differential trials of high and low-concentration odor exposures. The line represents the regression fit between the behavioral and neural responses (hex (L), R2 = 0.65; hex (H), R2 = 0.79). The negative slope of the regression line indicates that while the neural responses diminish over trials, the behavioral responses increase over trials.

(b-d) Similar plots as in panel a but for oct (L/H), iaa (L/H), and bza (L/H).

Neural and behavioral changes due to adaptation are odor specific

Finally, we wondered if the neural response reduction and behavioral enhancements were a global, non-specific state change in the olfactory system brought about by the repetition of any odorant, or are the observed neural and behavioral response changes odor specific. To examine this, we used a ‘catch-trial paradigm.’ (Fig. 7a) In these sets of experiments, we repeatedly presented an odorant to induce adaptation. After a substantial reduction in the neural response or stabilization of the behavioral response, a deviant or catch stimulus was presented. Persistence of the observed reduction in neural responses and enhancement in behavioral responses when the catch stimulus is introduced would reveal that the adaptation-induced changes are not odor specific.

Adaptation results in odor-specific neural response reductions and behavioral output facilitation.

(a)Two blocks of trials were used. First, a block of 30 trials where one odorant (hex) was presented in all trials except the 26th trial (the catch trial). During the catch trial, a deviant stimulus (iaa or app) was presented. After a 15-minute no-odor reset period, a second block of ten trials of the deviant stimulus was presented (trials 31 – 40). This was done to determine the unadapted (first trial) and adapted (later trials) responses of the same set of PNs to the stimulus used in the catch trial.

(b)Summed spike counts (during 4 s odor presentation period) across all PNs were calculated and shown as a function of trials (hex repetitions). Repeated presentations of odor A resulted in a substantial reduction in odor-evoked spike counts. However, the presentation of the deviant odor (iaa) during the catch trial (trial # 26) resulted in a marked increase in spike counts. The response strength was similar to the non-adapted responses (Trial # 31) for iaa (H). (right) Similar results are shown when hex was repeatedly presented and app was used as the deviant stimulus. Note that the response to app during the catch trial did not recover as was noted for iaa (left).

(c)Correlations between odor-evoked responses with hex response in the first trial are shown as a function of trial number.

(d)A similar catch-trial paradigm was followed for behavioral experiments. Each locust was presented with a repeating odorant for eight trials. A deviant stimulus was presented in the ninth trial. This was followed by six more repetitions of the recurring stimulus.

(e)Response matrix summarizing POR responses of individual locust PORs are shown. Same convention as in Fig. 1. (f)The P(POR) across locusts was calculated and plotted as a function of the trial number. Notably, when the deviant stimulus was presented there is a marked decrease in P(POR). The dotted line indicates the unadapted P(POR) value for the deviant stimulus.

We used the catch-trial experiments with two different catch stimuli. Our neural recordings reveal that when iaa (ester) was used as the catch stimulus after repeated presentations of hex (alcohol), the neural responses strength recovered and increased to an overall higher-spike count (Fig. 7b). The observed catchiaa response strength matched the unadapted response level to iaa when it was separately presented following a 15-minute reset window (i.e. Fig. 7b, trial# 31). In contrast, when apple (a complex mixture) was used as the catch stimulus, the neural responses did not recover to higher levels indicating significant cross-adaptation (Fig. 7b). Nevertheless, in both cases, the ensemble responses evoked by the catch stimulus were highly similar to the unadapted responses it usually generates (Fig. 7c). Therefore, these results indicate that there is stimulus specificity to the adaptation-induced changes in neural responses and that cross-adaptation does not corrupt the identity of the deviant stimulus.

Behaviorally, we chose odor pairs where the component odorants had markedly different levels of POR responses (Fig. 7d). Repeated presentations of an odorant (bza or iaa) increased the P(POR) across the locusts as observed before (Fig. 7e). Notably when the catch stimulus was presented, the behavioral 1responses dropped close to the first-trial P(POR) elicited by that odorant (Fig. 7f; dotted line). These findings were also observed when the recurring and catch odorants were switched (Fig 7f).

In sum, these neural and behavioral results indicate that the observed neural and behavioral response changes brought about by the repeated encounters with an odorant are at least partially odorspecific and are not global non-specific changes in the olfactory system.

Discussion

Sensory systems often have to deal with constraints that are of an opposing nature. For example, the ability to maintain stable representations is key to robust recognition of sensory stimuli42–44. Often variations in the external environment, such as humidity and temperature, may introduce changes in neural responses that have to be filtered out to preserve the identity of the stimuli. Additionally, the strength or intensity of the stimulus can also change widely. Compressing or even removing these variations might be necessary to recognize an odorant independent of its intensity42–47. Alternately, in certain cases, it might also become important to maintain some of these variations so they can be exploited for guiding or altering the behavioral response. As an example, it might be worth noting that many chemicals that are pleasant at lower intensities can become repelling or even harmful to the organism at extremely high intensities13–15. Hence, the behavioral responses should accordingly vary with intensity. Further, many organisms have shown that they can follow an odorant to their source (source localization). Such behaviors are of ecological importance as they allow the organisms to find the food source or potential mates. In all these cases, it becomes important to maintain variations with stimulus intensity to ensure the organisms are able to approach the odor source. In our earlier studies, we have focused on examining how robust odor recognition is achieved37,40,48. Here we examined the latter problem of how to maintain information about the stimulus intensity while the responses in the neural circuits are becoming weaker due to adaptation.

Consistent with earlier studies30,49,50, our data revealed that the response of individual projection neurons in the antennal lobe varied with the repetition of the same stimulus. While a reduction in odorevoked responses was observed in many neurons, an increase in spiking activity and changes in temporal firing patterns was also observed. Notably, these changes were not random but highly repeatable (Fig. 2c). Hence, these results indicate that the effect of adaptation in a neural circuit is highly consistent and reliable.

In addition to spike rate reduction, oscillatory synchronization, and inter-neuronal coherence build up in the locust antennal lobe over repeated encounters with an odorant30,51,52. This short-term memory was not observed in the antenna but endured in the antennal lobe for several minutes after the termination of the stimulus. Similar results were also reported in the mouse olfactory bulb43. The inputs from sensory neurons were shown to be consistent across repetition whereas the outputs of the olfactory bulb were systematically reduced with stimulus repetition. These results suggest that loci of short-term adaptation, at least when the stimulus is discontinuous and recurring, is in the antennal lobe/olfactory bulb circuitry. Although the precise mechanism for achieving the same is not fully understood49,53.

What is the computational significance of this neural adaptation? Sensory adaptation has been implicated in several important computations such as high-pass filtering, matching neural response to stimulus statistics, generating sparser codes, and optimal representations33,35,54–56. In the olfactory system, adaptation has also been suggested to stabilize neural representation such that recognition of odor identity from ensemble neural responses improved after repeated encounters with the stimulus30. However, adaptation also introduces a potential confound whereby the absolute neural response strength becomes an unreliable indicator of stimulus strength. Our results extend these earlier studies by revealing how this potential confound could be resolved in the olfactory system.

Increases in stimulus intensity have also been shown to result in higher response amplitudes in firing neurons and in the recruitment of additional activated neurons35,42,44,57. While our results show that this is indeed true, we also found that several neurons responded preferentially during the lower concentration exposures but did not respond when the higher intensity of the same stimulus was presented (Supp. Fig. 2). Hence, the activated combination of neurons differed both with stimulus identity and intensity. Furthermore, even though adaptation reduced the overall spiking activities across neurons, the combination of neurons activated was robustly maintained. In other words, the population response vector direction (determined by the combination of neurons activated) was consistent. Only the vector length (determined by the total spikes across neurons) is reduced with the repetition of the stimulus. This simple approach was sufficient to encode the information about the odor intensity in an adaptation-invariant fashion. Additionally, since equivalent information about stimulus identity and intensity was preserved with fewer spikes, neural adaptation can be considered as a method of optimizing the neural representation2,3,9,10,33,34.

Does the observed persistence of concentration information in the neural activities after adaptation translate to observable differences in behavioral outcomes? Using an innate palp-opening response (POR) we examined how behavioral responses varied with stimulus repetition. Unlike the neural responses, we found that the PORs increased with repetition and after about five trials settled into odor-specific levels of responses. Notably, the PORs in the later trials were significantly higher for higher stimulus intensities for most of the odorants used. In sum, our results showed that locusts are better able to behaviorally differentiate stimulus intensity after adaptation.

Alternately, adaptation due to the repetition of an odorant could alter the perceived intensity of the recurring odorant or other related test odorants. The compromise of stimulus intensity information due to adaptation has been observed in behavioral discrimination tasks in rats and even in psychophysical studies in humans58,59. Such a result can arise when the combinatorial response patterns evoked by the adapting and the test stimulus have significant overlap. However, our results indicate that when changes in odor intensities alter the combination of neurons activated, it is feasible to maintain information about odor intensity levels in a robust fashion.

Finally, we note that the anticorrelated link between neural suppression and behavioral enhancements (called ‘repetition priming’) is well established in other animal models60. How can weaker neural responses lead to enhanced behavioral outcomes? Our results indicate that the antennal lobe neural responses in later trials were not sharpened by pruning out weaker responses (Supp. Fig. 2). The weakening of the response strength could potentially lead to a sparser response in the downstream mushroom body. In addition, as we noted earlier, neural response synchrony increases with stimulus repetition. Whether weaker but synchronized activity in neurons drives sparser and more selective responses which then produce enhanced behavioral outcomes remains to be carefully examined.

eLife Assessment

This study addresses an important question in sensory neuroscience: how the olfactory system distinguishes decreases in stimulus intensity from decreases in neural responses due to adaptation. Based on a combination of electrophysiological and behavioral analyses, solid evidence establishes that neural coding changes differently between intensity reductions and adaptation. Intriguingly, behavioral responses tend to increase as the neural responses decrease, suggesting that core features of the odor response persist through adaptation. While the experimental results are convincing overall, the conclusions would be strengthened by more refined statistical analysis and data quantification.

Reviewer #1 (Public Review):

The authors use electrophysiological and behavioral measurements to examine how animals could reliably determine odor intensity/concentration across repeated experiences. Because stimulus repetition leads to short-term adaptation evidenced by reduced overall firing rates in the antennal lobe and firing rates are otherwise concentration-dependent, there could be an ambiguity in sensory coding between reduced concentration or more recent experience. This would have a negative impact on the animal's ability to generate adaptive behavioral responses that depend on odor intensities. The authors conclude that changes in concentration alter the constituent neurons contributing to the neural population response, whereas adaptation maintains the 'activated ensemble' but with scaled firing rates. This provides a neural coding account of the ability to distinguish odor concentrations even after extended experience. Additional analyses attempt to distinguish hypothesized circuit mechanisms for adaptation but are inconclusive. A larger point that runs through the manuscript is that overall spiking activity has an inconsistent relationship with behavior and that the structure of population activity may be the more appropriate feature to consider.

To my knowledge, the dissociation of effects of odor concentration and adaptation on olfactory system population codes was not previously demonstrated. This is a significant contribution that improves on any simple model based on overall spiking activity. The primary result is most strikingly supported by visualization of a principal components analysis in Figure 4. However, there are some weaknesses in the data and analyses that limit confidence in the overall conclusions.

1. Behavioral work interpreted to demonstrate discrimination of different odor concentrations yields inconsistent results. Only two of the four odorants follow the pattern that is emphasized in the text (Figure 1F). Though it's a priori unlikely that animals are incapable of distinguishing odor concentrations at any stage in adaptation, the evidence presented is not sufficient to reach this conclusion.

2. While conclusions center on concepts related to the combination of activated neurons or the "active ensemble", this specific level of description is not directly demonstrated in any part of the results. We see individual neural responses and dimensional reduction analyses, but we are unable to assess to what extent the activated ensemble is maintained across experience.

3. There is little information about the variance or statistical strength of results described at the population level. While the PCA presents a compelling picture, the central point that concentration changes and adaptation alter population responses across separable dimensions is not demonstrated quantitatively. The correlation analysis that might partially address this question is presented to be visually interpreted with no additional testing.

4. Results are often presented separately for each odor stimulus or for separate datasets including two odor stimuli. An effort should be made to characterize patterns of results across all odor stimuli and their statistical reliability. This concern arises throughout all data presentations.

5. The relevance of the inconclusive analysis of inferred adaptation mechanisms in Figure 2d-f and the single experiment including a complex mixture in Figure 7 to the motivating questions for this study are unclear.

6. Throughout the description of the results, typical standards for statistical reporting (sample size, error bars, etc.) are not followed. This prevents readers from assessing effect sizes and undermines the ability to assign a confidence to any particular conclusion.

Reviewer #2 (Public Review):

Summary:

The authors' main goal was to evaluate how both behavioral responses to odor, and their early sensory representations are modified by repeated exposure to odor, asking whether the process of adaptation is equivalent to reducing the concentration of an odor. They open with behavioral experiments that actually establish that repeated odor presentation increases the likelihood of evoking a behavioral response in their experimental subjects - locusts. They then examine neural activity patterns at the second layer of the olfactory circuit. At the population level, repeated odor exposure reduces total spike counts, but at the level of individual cells there seems to be no consistent guiding principle that describes the adaptation-related changes, and therefore no single mechanism could be identified.

Both population vector analysis and pattern correlation analysis indicate that odor intensity information is preserved through the adaptation process. They make the closely related point that responses to an odor in the adapted state are distinct from responses to lower concentration of the same odor. These analyses are appropriate, but the point could be strengthened by explicitly using some type of classification analysis to quantify the adaptation effects. e.g. a confusion matrix might show if there is a gradual shift in odor representations, or whether there are trials where representations change abruptly.

Strengths:

One strength is that the work has both behavioral read-out of odor perception and electrophysiological characterization of the sensory inputs and how both change over repeated stimulus presentations. It is particularly interesting that behavioral responses increase while neuronal responses generally decrease. Although the behavioral effect could occur fully downstream of the sensory responses the authors measure, at least those sensory responses retain the core features needed to drive behavior despite being highly adapted.

Weaknesses:

Ultimately no clear conceptual framework arises to understand how PN responses change during adaptation. Neither the mechanism (vesicle depletion versus changes in lateral inhibition) nor even a qualitative description of those changes. Perhaps this is because much of the analysis is focused on the entire population response, while perhaps different mechanisms operate on different cells making it difficult to understand things at the single PN level.

From the x-axis scale in Fig 2e,f it appeared to me that they do not observe many strong PN responses to these stimuli, everything being < 10 spikes/sec. So perhaps a clearer effect would be observed if they managed to find the stronger responding PNs than captured in this dataset.

Reviewer #3 (Public Review):

Summary:

How does the brain distinguish stimulus intensity reduction from response reductions due to adaptation? Ling et al study whether and how the locust olfactory system encodes stimulus intensity and repetition differently. They show that these stimulus manipulations have distinguishable effects on population dynamics.

Strengths:

1. Provides a potential strategy with which the brain can distinguish intensity decrease from adaptation. -- while both conditions reduce overall spike counts, intensity decrease can also changes which neurons are activated and adaptation only changes the response magnitude without changing the active ensemble.

2. By interleaving a non-repeated odor, they show that these changes are odor-specific and not a non-specific effect.

3. Describes how proboscis orientation response (POR) changes with stimulus repetition., Unlike the spike counts, POR increases in probability with stimulus. The data portray the variability across subjects in a clear way.

Weaknesses:

1. Behavior

a. While the "learning curve" of the POR is nicely described, the behavior itself receives very little description. What are the kinematics of the movement, and do these vary with repetition? Is the POR all-or-nothing or does it vary trial to trial?

b. What are the reaction times? This can constrain what time window is relevant in the neural responses. E.g., if the reaction time is 500 ms, then only the first 500 ms of the ensemble response deserves close scrutiny. Later spikes cannot contribute.

c. The behavioral methods are lacking some key information. While references are given to previous work, the reader should not be obligated to look at other papers to answer basic questions: how was the response measured? Video tracking? Hand scored?

d. Can we be sure that this is an odor response? Although airflow out of the olfactometer is ongoing throughout the experiment, opening and closing valves usually creates pressure jumps that are likely to activate mechanosensors in the antennae.

e. What is the baseline rate of PORs in the absence of stimuli?

e.What can you say about the purpose of the POR? I lack an intuition for why a fly would wiggle the maxillary palps. This is a question that is probably impossible to answer definitively, but even a speculative explanation would help the reader better understand.

2. Physiology

a. Does stimulus repetition affect "spontaneous" activity (i.e., firing in the interstimulus interval? To study this question, in Figures 2b and c, it would be valuable to display more of the pre-stimulus period, and a quantification of the stability or lability of the inter-stimulus activity.

b. When does the response change stabilize? While the authors compare repetition 1 to repetition 25, from the rasters it appears that the changes have largely stabilized after the 3rd or 4th repetition. In Figure 5, there is a clear difference between repetition 1-3 or so and the rest. Are successive repetitions more similar than more temporally-separated repetitions (e.g., is rep 13 more similar to 14 than to 17?). I was not able to judge this based on the dendrograms of Figure 5. If the responses do stabilize at it appears, it would be more informative to focus on the dynamics of the first few repetitions.

c. How do temporal dynamics change? Locust PNs have richly varied temporal dynamics, but how these may be affected is not clear. The across-population average is poorly suited to capture this feature of the activity. For example, the PNs often have an early transient response, and these appear to be timed differently across the population. These structures will be obscured in a cross population average. Looking at the rasters, it looks like the initial transient changes its timing (e.g., PN40 responses move earlier; PN33 responses move later.). Quantification of latency to first spike after stimulus may make a useful measure of the dynamics.

d. How legitimate is the link between POR and physiology? While their changes can show a nice correlation, the fact the data were taken from separate animals makes them less compelling than they would be otherwise. How feasible is it to capture POR and physiology in the same prep? This would be most helpful, but I suspect may be too technically challenging to be within scope.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public Review):

The authors use electrophysiological and behavioral measurements to examine how animals could reliably determine odor intensity/concentration across repeated experiences. Because stimulus repetition leads to short-term adaptation evidenced by reduced overall firing rates in the antennal lobe and firing rates are otherwise concentration-dependent, there could be an ambiguity in sensory coding between reduced concentration or more recent experience. This would have a negative impact on the animal's ability to generate adaptive behavioral responses that depend on odor intensities. The authors conclude that changes in concentration alter the constituent neurons contributing to the neural population response, whereas adaptation maintains the 'activated ensemble' but with scaled firing rates. This provides a neural coding account of the ability to distinguish odor concentrations even after extended experience. Additional analyses attempt to distinguish hypothesized circuit mechanisms for adaptation but are inconclusive. A larger point that runs through the manuscript is that overall spiking activity has an inconsistent relationship with behavior and that the structure of population activity may be the more appropriate feature to consider.

To my knowledge, the dissociation of effects of odor concentration and adaptation on olfactory system population codes was not previously demonstrated. This is a significant contribution that improves on any simple model based on overall spiking activity. The primary result is most strikingly supported by visualization of a principal components analysis in Figure 4. However, there are some weaknesses in the data and analyses that limit confidence in the overall conclusions.

We thank the reviewer for evaluating our work and highlighting its strengths and deficiencies. We have revised the manuscript with expanded behavioral datasets and additional analyses that we believe convincingly support our conclusion.

(1) Behavioral work interpreted to demonstrate discrimination of different odor concentrations yields inconsistent results. Only two of the four odorants follow the pattern that is emphasized in the text (Figure 1F). Though it's a priori unlikely that animals are incapable of distinguishing odor concentrations at any stage in adaptation, the evidence presented is not sufficient to reach this conclusion.

We have expanded our dataset and now show that the behavioral response is significantly different for high and low concentration exposures of the same odorant. This was observed for all four odorants in our study (refer to Revised Fig. 1F).

(2) While conclusions center on concepts related to the combination of activated neurons or the "active ensemble", this specific level of description is not directly demonstrated in any part of the results. We see individual neural responses and dimensional reduction analyses, but we are unable to assess to what extent the activated ensemble is maintained across experience.

We have done several additional analyses (see provisional response). Notably, we have corroborated our dimensionality reduction and correlation analysis results with a quantitative classification analysis that convincingly demonstrates that odor identity and intensity of the odorant can be decoded from the ensemble neural activity, and this could be achieved in an adaptation-invariant fashion (refer to Revised Supplementary Fig. 4).

(3) There is little information about the variance or statistical strength of results described at the population level. While the PCA presents a compelling picture, the central point that concentration changes and adaptation alter population responses across separable dimensions is not demonstrated quantitatively. The correlation analysis that might partially address this question is presented to be visually interpreted with no additional testing.

We have included a plot that compares the odor-evoked responses across all neurons (mean ± variance) at both intensity levels for each odorant (Revised Supplementary Fig. 5). This plot clearly shows how the ensemble neural activity profile varies with odor intensity and how these response patterns are robustly maintained across trials.

(4) Results are often presented separately for each odor stimulus or for separate datasets including two odor stimuli. An effort should be made to characterize patterns of results across all odor stimuli and their statistical reliability. This concern arises throughout all data presentations.

We had to incorporate a 15-minute window between presentations of odorants to reset adaptation. Due to this, we were unable to extracellularly record from all four odorants at two intensities from a single experiment (~ 3.5 hours of recording for just 2 odorants at two intensities with one odorant at higher intensity repeated at the end; Fig. 2a). Therefore, we recorded two datasets. Each dataset captured the responses of ~80 PNs to two odorants at two intensities, one odorant at the higher concentration repeated at the end of the experiment to show repeatability of changes due to adaptation.

(5) The relevance of the inconclusive analysis of inferred adaptation mechanisms in Figure 2d-f and the single experiment including a complex mixture in Figure 7 to the motivating questions for this study are unclear.

Figure 2d-f has been revised. While we agree that the adaptation mechanisms are not fully clear, there is a trend that the most active PNs are the neurons that change the most across trials. This change and the response in the first trial are negatively correlated, indicating that vesicle depletion could be an important contributor to the observed results. However, neurons that adapt strongly at higher intensities are not the ones that adapt at lower intensities. This complicates the understanding of how neural responses vary with intensities and the adaptation that happens due to repetition. This has been highlighted in the revised manuscript.

Regarding Figure 7, we wanted to examine the odor-specificity of the changes that happen due to repeated encounters of an odorant. Specifically, wondered if the neural response reduction and behavioral enhancements were a global, non-specific state change in the olfactory system brought about by the repetition of any odorant, or are the observed neural and behavioral response changes odor-specific.

(6) Throughout the description of the results, typical standards for statistical reporting (sample size, error bars, etc.) are not followed. This prevents readers from assessing effect sizes and undermines the ability to assign a confidence to any particular conclusion.

We have revised the manuscript to fix these issues and included sample size and error bars in our plots.

Reviewer #2 (Public Review):

Summary:

The authors' main goal was to evaluate how both behavioral responses to odor, and their early sensory representations are modified by repeated exposure to odor, asking whether the process of adaptation is equivalent to reducing the concentration of an odor. They open with behavioral experiments that actually establish that repeated odor presentation increases the likelihood of evoking a behavioral response in their experimental subjects - locusts. They then examine neural activity patterns at the second layer of the olfactory circuit. At the population level, repeated odor exposure reduces total spike counts, but at the level of individual cells there seems to be no consistent guiding principle that describes the adaptation-related changes, and therefore no single mechanism could be identified.

Both population vector analysis and pattern correlation analysis indicate that odor intensity information is preserved through the adaptation process. They make the closely related point that responses to an odor in the adapted state are distinct from responses to lower concentration of the same odor. These analyses are appropriate, but the point could be strengthened by explicitly using some type of classification analysis to quantify the adaptation effects. e.g. a confusion matrix might show if there is a gradual shift in odor representations, or whether there are trials where representations change abruptly.

Strengths:

One strength is that the work has both behavioral read-out of odor perception and electrophysiological characterization of the sensory inputs and how both change over repeated stimulus presentations. It is particularly interesting that behavioral responses increase while neuronal responses generally decrease. Although the behavioral effect could occur fully downstream of the sensory responses the authors measure, at least those sensory responses retain the core features needed to drive behavior despite being highly adapted.

Weaknesses:

Ultimately no clear conceptual framework arises to understand how PN responses change during adaptation. Neither the mechanism (vesicle depletion versus changes in lateral inhibition) nor even a qualitative description of those changes. Perhaps this is because much of the analysis is focused on the entire population response, while perhaps different mechanisms operate on different cells making it difficult to understand things at the single PN level.

From the x-axis scale in Fig 2e,f it appeared to me that they do not observe many strong PN responses to these stimuli, everything being < 10 spikes/sec. So perhaps a clearer effect would be observed if they managed to find the stronger responding PNs than captured in this dataset.

We thank the reviewer for his/her evaluation of our work. Indeed, our work does not clarify the mechanism that underlies the adaptation over trials, and how this mechanism accounts for adaptation that is observed at two different intensities of the same odorant. However, as we highlight in the revised manuscript, there is some evidence for the vesicle depletion hypothesis. For the plots shown in Fig. 2, the firing rates were calculated after averaging across time bins and trials. Hence, the lower firing rates. The peak firing rates of the most active neurons are ~100 Hz. So, we are certain that we are collecting responses from a representative ensemble of neurons in this circuit.

Reviewer #3 (Public Review):

Summary:

How does the brain distinguish stimulus intensity reduction from response reductions due to adaptation? Ling et al study whether and how the locust olfactory system encodes stimulus intensity and repetition differently. They show that these stimulus manipulations have distinguishable effects on population dynamics.

Strengths:

(1) Provides a potential strategy with which the brain can distinguish intensity decrease from adaptation. -- while both conditions reduce overall spike counts, intensity decrease can also changes which neurons are activated and adaptation only changes the response magnitude without changing the active ensemble.

(2) By interleaving a non-repeated odor, they show that these changes are odor-specific and not a non-specific effect.

(3) Describes how proboscis orientation response (POR) changes with stimulus repetition., Unlike the spike counts, POR increases in probability with stimulus. The data portray the variability across subjects in a clear way.

We thank the reviewer for the summary and for highlighting the strengths of our work.

Weaknesses:

(1) Behavior

a. While the "learning curve" of the POR is nicely described, the behavior itself receives very little description. What are the kinematics of the movement, and do these vary with repetition? Is the POR all-or-nothing or does it vary trial to trial?

The behavioral responses were monitored in unconditioned/untrained locusts. Hence, these are innate responses to the odorants. These innate responses are usually brief and occur after the onset of the stimulus. However, there is variability across locusts and trials (refer Revised Supplementary Fig. 1). When the same odorant is conditioned with food reward, the POR responses become more stereotyped and occur rapidly within a few hundred milliseconds.

POR response dynamics in a conditioned locust.

The palps were painted in this case (left panel), and the distance between the palps was tracked as a function of time (right panel).

b. What are the reaction times? This can constrain what time window is relevant in the neural responses. E.g., if the reaction time is 500 ms, then only the first 500 ms of the ensemble response deserves close scrutiny. Later spikes cannot contribute.

This is an interesting point. We had done this analysis for conditioned POR responses. For innate POR, as we noted earlier, there is variability across locusts. Many responses occur rapidly after odor onset (<1 s), while some responses do occur later during odor presentation and in some cases after odor termination. It is important to note that these dynamical aspects of the POR response, while super interesting, should occur at a much faster time scale compared to the adaptation that we are reporting across trials or repeated encounters of an odorant.

c. The behavioral methods are lacking some key information. While references are given to previous work, the reader should not be obligated to look at other papers to answer basic questions: how was the response measured? Video tracking? Hand scored?

We agree and apologize for the oversight. We have revised the methods and added a video to show the POR responses. Videos were hand-scored.

d. Can we be sure that this is an odor response? Although airflow out of the olfactometer is ongoing throughout the experiment, opening and closing valves usually creates pressure jumps that are likely to activate mechanosensors in the antennae.

Interesting. We have added a new Supplementary Fig. 2 that shows that the POR to even presentations of paraffin oil (solvent; control) is negligible. This should confirm that the POR is a behavioral response to the odorant.

Furthermore, all other potential confounds identified by the reviewer are present for every odorant and every concentration presented. However, the POR varies in an odor-identity and intensity-specific manner.

e. What is the baseline rate of PORs in the absence of stimuli?

Almost zero.

f. What can you say about the purpose of the POR? I lack an intuition for why a fly would wiggle the maxillary palps. This is a question that is probably impossible to answer definitively, but even a speculative explanation would help the reader better understand.

The locusts use these finger-like maxillary palps to grab a grass blade while eating. Hence, we believe that this might be a preparatory response to feeding. We have noted that the PORs are elicited more by food-related odorants. Hence, we think it is a measure of odor appetitiveness. This has been added to the manuscript.

(2) Physiology

a. Does stimulus repetition affect "spontaneous" activity (i.e., firing in the interstimulus interval? To study this question, in Figures 2b and c, it would be valuable to display more of the prestimulus period, and a quantification of the stability or lability of the inter-stimulus activity.

Done. Yes, the spontaneous activity does appear to change in an odor-specific manner. We have done some detailed analysis of the same in this preprint:

Ling D, Moss EH, Smith CL, Kroeger R, Reimer J, Raman B, Arenkiel BR. Conserved neural dynamics and computations across species in olfaction. bioRxiv [Preprint]. 2023 Apr 24:2023.04.24.538157. doi: 10.1101/2023.04.24.538157. PMID: 37162844; PMCID: PMC10168254

b. When does the response change stabilize? While the authors compare repetition 1 to repetition 25, from the rasters it appears that the changes have largely stabilized after the 3rd or 4th repetition. In Figure 5, there is a clear difference between repetition 1-3 or so and the rest. Are successive repetitions more similar than more temporally-separated repetitions (e.g., is rep 13 more similar to 14 than to 17?). I was not able to judge this based on the dendrograms of Figure 5. If the responses do stabilize at it appears, it would be more informative to focus on the dynamics of the first few repetitions.

The reviewer makes an astute observation. Yes, the changes in firing rates are larger in the first three trials (Fig. 3c). The ensemble activity patterns, though, are relatively stable across all trials as indicated by the PCA plots and classification analysis results.

Correlation as a function of trial number.

All correlations were made with respect to the odor-evoked responses in the last odor trial of hex(H) and bza(H).

c. How do temporal dynamics change? Locust PNs have richly varied temporal dynamics, but how these may be affected is not clear. The across-population average is poorly suited to capture this feature of the activity. For example, the PNs often have an early transient response, and these appear to be timed differently across the population. These structures will be obscured in a cross population average. Looking at the rasters, it looks like the initial transient changes its timing (e.g., PN40 responses move earlier; PN33 responses move later.). Quantification of latency to first spike after stimulus may make a useful measure of the dynamics.

As noted earlier, to keep our story simple in this manuscript, we have only focused on the variations across trials (i.e., much slower response dynamics). We did this as we are not recording neural and behavioral responses from the same locust. We plan to do this and directly compare the neural and behavioral dynamics in the same locust.

d.How legitimate is the link between POR and physiology? While their changes can show a nice correlation, the fact the data were taken from separate animals makes them less compelling than they would be otherwise. How feasible is it to capture POR and physiology in the same prep?

This would be most helpful, but I suspect may be too technically challenging to be within scope.

The antennal lobe activity in the input about the volatile chemicals encountered by the locust. The POR is a behavioral output. Hence, we believe that examining the correlation between the olfactory system's input and output is a valid approach. However, we have only compared the mean trends in neural and behavioral datasets, and dynamics on a much slower timescale. We are currently developing the capability to record neural responses in behaving animals. This turned out to be a bit more challenging than we had envisioned. We plan to do fine-grained comparisons of the neural and behavioral dynamics, recommended by this reviewer, in those preparations.

Further, we will also be able to examine whether the variability in behavioral responses could be predicted from neural activity changes in that prep.

Version 2

Abstract

Abstract

Neural responses evoked by a stimulus reduce upon repetition. While this adaptation allows the sensory system to attend to novel cues, does information about the recurring stimulus, particularly its intensity, get compromised? We explored this issue in the locust olfactory system. We found that locusts’ innate behavioral response to odorants varied with repetition and stimulus intensity. Counter-intuitively, the stimulus-intensity dependent differences became significant only after adaptation had set in. Adaptation altered responses of individual neurons in the antennal lobe (neural network downstream to insect antenna). These response variations to repetitions of the same stimulus were unpredictable and inconsistent across intensities. Although both adaptation and intensity decrements resulted in an overall reduction in spiking activities across neurons, these changes could be disentangled, and information about stimulus intensity was robustly maintained by ensemble neural responses. In sum, these results show how information about odor intensity can be preserved in an adaptation-invariant manner.

Introduction

The ability to adapt to sensory cues is key to the survival of many living organisms1–8. The goal of this sensory adaptation is to condense neural responses to a persisting or recurring stimulus in order to focus on novel cues. Such condensed neural responses are thought to increase system efficiency by reducing the metabolic requirements of information encoding9–12. However, is sensory information lost when neural responses are condensed during adaptation? While this computational task appears relatively elementary, any solution should satisfy two important requirements. First, attenuation of stimulus-evoked responses due to adaptation should not alter the identity of the stimulus. Second, information regarding the intensity of the stimulus may have to be preserved when odor intensity conveys information of behavioral importance. The latter requirement is paramount when sensory stimuli have different behavioral significance at varying intensities. For example, some odorants are attractive at lower intensities but switch to being aversive beyond certain threshold values13–15. Sensitivity to intensity gradients has to be maintained for sustained periods to allow for navigation towards or away from the stimulus source16–18. Such biological contexts necessitate the preservation of odor intensity information across repeated encounters with the odorant. However, the mechanism by which information about stimulus intensity is robustly maintained is not fully understood.

Sensory adaptation occurs over both long- and short-time scales12,19–21. Long-term adaptation requires gene transcription and protein translation that occurs on the time scale of hours to days6,22–24, whereas short-term adaptation occurs on a millisecond to minute time scale2,11,25. Short-term adaptation results from a wide range of mechanisms with prominent models in the olfactory system being vesicle depletion and facilitation of lateral inhibition25–32. While both vesicle depletion and lateral inhibition are likely to occur in the olfactory system, these complementary mechanisms will manifest in two different, but testable, outcomes. Vesicle depletion will impact the most prominent neurons with elevated odor-evoked responses, whereas facilitation of lateral inhibition would progressively suppress activity in the relatively weak responders. Irrespective of which mechanisms underlie olfactory adaptation, the neural responses evoked by a stimulus will change over time and across repeated encounters. This raises the question: Does adaptation then corrupt or alter information regarding the odorant?

The ambiguity that neural adaptation introduces has been documented in visual33 and auditory systems34,35. Prior studies have shown that odor identity is robustly maintained by the spatiotemporally patterned neural responses and can allow precise recognition across repeated encounters with the odorant30,36–38. But is intensity information corrupted by the changes in neural activity due to sensory adaptation? Both reductions in stimulus intensity and sensory adaptation lead to a reduction in the overall spiking responses in most neural circuits. Can these neural response variations due to neural adaptation and intensity decrements be disentangled? We sought to examine this in the relatively simple insect olfactory system.

Our results show that odor intensity changes result in subtle variations in the overall combination of activated neurons. While sensory adaptation results in an overall reduction in the population neural responses, the activated ensemble is robustly maintained to allow simultaneous and precise decoding of odor identity and intensity. Using an innate behavioral assay, we show that physiological adaptation coincided with enhanced behavioral differentiation between odorant intensities. Thus, our work reveals a simple but elegant approach to achieving adaptation-invariant intensity coding and recognition in the insect olfactory system.

Results

Behavioral responses to an odorant vary with repetition and intensity

We began by examining how behavioral responses change with repetitions of an odorant. We used an innate behavioral assay where the presentation of certain odorants triggered an appetitive palp-opening response (POR) in starved locusts (Fig. 1a; Supp. Fig. 1). Typically, these palps are used to grab food such as a blade of grass and can also be triggered when locusts encounter certain food-related odorants. Hence, it has been used as a measure of the appetitiveness of an odorant39–41. For this study, we regarded the probability of an odor eliciting a POR (p(POR)) as an indicator for stimulus perception. The baseline p(POR), in the absence of any olfactory stimuli, was negligible. Further, presentations of puffs of paraffin oil (control) evoked only weak palp-opening responses (Supp. Fig. 2). Note that the p(POR) was computed by averaging the behavioral responses across locusts for each trial or repetition number (Fig. 1b–d). Intriguingly, we found that the p(POR) systematically increased with repetitions of the same odorant and stabilized after about five stimulus presentations.

Behavioral response facilitation upon odor repetition.

(a) Schematic of the innate palp-opening response (POR) experimental paradigm. Opening of the locust’s maxillary palps within 15 seconds of the odor onset was considered a POR.

(b) Schematic of the block odor stimulation protocol. Each block consisted of ten trials, and a 4s odor pulse was presented in each trial. A 15-minute no-stimulation period separated the blocks.

(c) Response matrices are shown summarizing individual locust PORs (rows) for blocks of ten trials (columns). White boxes indicate PORs in a specific trial, while black boxes indicate the absence of PORs in that trial. Locusts were sorted such that the least responsive locusts are shown at the top and the most responsive ones are near the bottom. PORs varied between locusts across trials.

(d) The probability of PORs across locusts (p(POR)) is shown as a function of trial number for two odorants: iaa and bza at high intensities (1% v/v).

(e) p(POR)s as a function of trial number is shown for four odorants (oct, hex, iaa, bza) at two different intensities (0.1% v/v – low and 1% v/v – high). The p(POR) of both odorant intensities increased over trials. Further, note that the p(POR) values for high-intensity odor exposures were notably higher than p(PORs) for low-intensity odor presentations.

(f) Bar plots compare the average P(POR) for the low- and high-intensity presentations of the same odor. p(POR) values observed at high and low intensities of an odorant were significantly different for all four odorants tested (p<0.05). n = 36 locusts for iaa, n = 33 locusts for hex, n = 47 locusts for oct, and n = 64 for bza.

Previous results have shown that the POR is a function of both stimulus identity and intensity39–41. Therefore, we presented each locust with randomized blocks of the same odorant at either high (H, 1% v/v) or low (L, 0.1% v/v) intensities (Fig. 1e). Our results indicate that the p(PORs) evoked by high and low stimulus intensities of an odorant were significantly different for all four odorants in our panel. (Fig. 1f; p ≤ 0.05; one-sided t-test). Notably, the behavioral responses to high and low concentration exposures of the same odorant were not significantly different in the earlier trials (Supp. Fig. 3; p > 0.05 for hex, iaa, and oct; one-sided t-test). However, the p(POR) became significantly different during the later trials (p ≤ 0.05 for all four odorants; one-sided t-test; trials six through ten). In other words, high and low stimulus intensities were more behaviorally discriminable in the later trials. Hence, these results counterintuitively suggest that behavioral responses to an odorant at two different intensities were different after multiple encounters with the same stimulus.

Odor-evoked response adaptation leads to complex changes in individual neurons

How do neural responses change as a function of stimulus repetition? To understand this, we examined the odor-evoked responses of individual projection neurons (PNs) in the locust antennal lobe as different odorants at two different intensities were repeatedly presented (Fig. 2a). Odor-evoked responses in individual PNs varied subtly across trials (Fig. 2b). In general, most PN responses to the odorant were stronger in the first trial and reduced during later trials of the same stimulus. Notably, the changes in PN responses in two different blocks of trials, where the same odorant was presented, showed repeatable changes in odor-evoked spiking patterns and the overall response strength (Fig. 2c).

Ensemble neural activity systematically changes across repeated encounters with an odorant.

(a) Schematic of the olfactory stimulation protocol. Each block consisted of twenty-five trials with a four-second odor pulse delivered in each trial. The inter-trial interval was 60 s. Two datasets were collected. Each dataset consisted of five randomized blocks of four odorants (dataset 1: hex (H), hex (L), oct (H), oct (L), hex (H)-repeat; dataset 2: iaa (H), iaa (L), bza (H), bza (L), iaa (H)-repeat). A fifteen-minute no-odor stimulation period separated blocks of trials.

(b) Raster plots of eight representative projection neurons (PNs) in the locust antennal lobe are shown. Spiking activities are shown for twenty-five trials (rows) with earlier trials shown at the top and later trials at the bottom. The shaded region indicates the four-second odor stimulation period, and the identity of the stimulus is indicated in each plot.

(c) Raster plots are shown for four representative PNs during two blocks of trials. The same odorant was presented in both blocks. Note that spiking activity changes are repeatable across the block of trials.

(d) Schematic showing how vesicular depletion and lateral inhibition facilitation models would change spiking activity in individual neurons. The Y-axis represents the change in response over trials (25th trial response – 1st trial response). Positive numbers indicate that the last trial had a stronger response, hence the response increase. Negative numbers indicate that the first trial had a stronger response, hence response reduction. Along the x-axis, the response in the first trial is shown. Vesicular depletion should impact the strongly activated neurons more, whereas lateral inhibition facilitation should progressively suppress weak responders.

(e) Following the schematic in panel d, response changes observed in PNs are shown. All PNs were included in this plot. Different colors and symbols are used to denote the identity of the odorant and its intensity (low-intensity trials – triangles and high-intensity trials – circles). As can be noted, the R2 values are∼ 0.48-0.56 indicating that the vesicle depletion facilitation model captures the adaptation trends in our datasets to a certain extent.

(f) Changes in PN spiking activity over trials during high-intensity odor exposures (y-axis) are plotted against response changes observed for the same PN during low-intensity exposures of the same odorant. The poor correlation values (hex, R2=0.17; oct, R2=0.12; iaa R2=0.21; bza R2=0.08) indicate that reductions in neural response amplitude for one odor intensity do not model reductions in response amplitude for another odor intensity.

Next, we examined how individual PNs adapted during stimulus repetitions. Based on existing literature, two widely considered models of adaptation in the olfactory system are vesicular depletion and facilitation of lateral inhibition. While the vesicular depletion model predicts that the strongest firing neurons would adapt the most, in the lateral inhibition facilitation model the weakly activated neurons would progressively be suppressed more (Fig. 2d). To test these two opposing but complementary theories, we calculated the total change in the total spike counts observed during the 25th and first odor presentations (Fig. 2e, y-axis; positive values indicate stronger response in the later trial and hence response facilitation, whereas negative values indicate reduction of odor-evoked responses). We plotted these response changes against the spike response observed in the first trial (Fig. 2e, x-axis). If vesicular depletion was the predominant form of adaptation, there would be a negative correlation with the response strength in the first trial. Whereas a positive correlation would indicate that lateral inhibition was the predominant form of adaptation. As can be noted, across both datasets with two concentrations of the four odorants tested, the vesicle depletion hypothesis appeared to capture some of the adaptation dynamics in our datasets (R2 = 0.48, 0.56, respectively).

Are trial-to-trial changes in stimulus-evoked neural responses at two different concentrations of the same odorant correlated? If this is the case, then it would be reasonable to expect neurons that adapt more at the higher concentration to also change in a similar fashion at a lower concentration and vice versa. To assess this, we calculated and plotted the change in spike counts for each PN at both high and low concentrations of the same odorant (Fig. 2f). Notably, the PNs with the greatest response change during repetitions of the higher-intensity odor exposures were not the ones that adapted heavily when encountering the same stimulus at a lower intensity (R2 = 0.17 (hex), 0.12 (oct), 0.21 (iaa), 0.08 (bza)). In sum, these results indicate that change due to adaptation in individual neurons is not a simple function of their response strength. Further, how individual neurons adapt is a function of both odor identity and intensity.

Adaptation and intensity decrements both reduce ensemble neural response strength

What is the overall change in the spiking response across the neural ensemble? To examine this, we calculated the average spiking response across all PNs recorded and plotted the spike counts as a function of time for each trial (Fig. 3a). Expectedly, the overall response strength was stronger in the first trial and reduced when the same stimulus was repeatedly encountered. A similar reduction in overall response strengths was also observed when the odorant intensity was reduced (Fig. 3b)

Stimulus repetition and intensity decrements reduce spiking responses.

(a) Peristimulus time histograms (PSTHs) across all PNs (hex and oct, n = 80 PNs; iaa and bza, n = 81 PNs) for each odorant. PSTHs of trials 1 and 25 are shown.

(b) PSTHs across all PNs are compared between high and low-intensity odor exposures. Response during the first trial is shown.

(c) Summed spike counts (across PNs and four-second odor presentation) are calculated and shown as a function of the trial number. The dotted line indicates the spike count of the 25th trial of the high-intensity odor exposure.

Are the response reductions during stimulus repetition comparable to those observed with stimulus intensity decrements? To determine this, we plotted the total spike counts summed across all PNs over the entire stimulus presentation duration and plotted it as a function of the trial number (Fig. 3c). We compared the changes over trials with the total spike counts observed during the high (H) and low (L) intensities of the same odorant. Notably, the total PN ensemble spike counts decreased due to both stimulus intensity reduction and repetition of the stimulus. Do these results indicate that adaptation could potentially confound information regarding stimulus intensity?

Adaptation-invariant encoding of odor intensity

Could the information regarding stimulus intensity be robustly encoded in the population neural responses? As noted earlier, both spiking activities in individual neurons and the volume of spikes at the ensemble level vary across trials. If adaptation alters the overall spiking activity amplitude (i.e., vector length; Fig. 4a), while changes in stimulus intensity alter the combination of activated neurons (i.e., vector direction; Fig. 4b), then information about odor identity and intensity can be preserved in an adaptation-invariant fashion. To test whether the recorded ensemble of neurons in our datasets encoded information in such an adaptation-invariant fashion, we first visualized the high-dimensional neural activities in each trial using a dimensionality reduction approach (see Methods). We plotted the ensemble responses in each 50 ms time bin during the odor presentation time window (4 s) and linked them based on the order of their occurrence to generate trial-by-trial odor response trajectories (Fig. 4c, d). Note that each trial generated a single loop response trajectory after dimensionality reduction. Further, neural responses in different trials evoked closed-loop trajectories that systematically reduced in length with increasing repetition but maintained their direction, indicating that the combination of neurons activated was mostly conserved. Different odorants evoked responses that were markedly different in the combination of activated neurons, and therefore, their response trajectories were oriented in different directions. Notably, different concentrations of the same odorant also generated response trajectories that subtly varied in direction. As a direct consequence, even though the overall spiking activities reduced over trials, the odor response trajectories reliably maintained their direction in all the trials. Hence, information about odor identity and intensity was robustly maintained across all trials.

Odor identity and intensity information are maintained across trials.

(a) A schematic showing how the ensemble neural activity might change between two odorants and across multiple trials or repetitions. The combination of neural activated should differ between different odorants, and therefore, the odor identity should be represented by population response vectors that differ in their direction. Repetitions should reduce response strength without altering the combination of neurons activated. If this were the case, the later repetitions of the same odorant would evoke a response that can be represented by vectors that maintain directions while progressively becoming shorter in length or magnitude.

(b) A schematic coding scheme for achieving adaptation invariant intensity coding. The combination of neurons activated changes markedly with odor identity and subtly with odor intensity. These responses become less intense without altering the combination of neurons activated. If this were the case, then the ensemble vector direction would change with both odor identity and intensity, and the vector direction robustly maintained even though the vector length continues to change with repetition.

(c) Trial-by-trial odor-evoked ensemble PN response trajectories after dimensionality-reduction are shown for hex and oct. Only high-concentration exposures of both odorants are included in this plot. For each odor, the temporal response trajectories for the 1st, 5th, 10th, 15th, 20th, and 25th trials are shown (color gradient from light to dark as trial number increases). Note that the odor-response trajectories become increasingly smaller during later trials, but the direction of the trajectories is robustly maintained.

(d) Trial-by-trial response trajectories for low-concentration exposures of the hex and oct are shown. Similar convention as in panel c.

(e) Similar plot as in panel c, but now comparing the odor-evoked response trajectories elicited by hex at high and low intensities.

(f) Similar plot as in panel d, but comparing the response trajectories elicited by oct at high and low intensities.

To quantify these observations, we performed a correlation analysis using high-dimensional PN response vectors. If a similar combination of PNs (high-dimensional activity vectors) were activated in two different trials, then their correlation would be high. Correlations across different trials and concentrations of the same odorant are shown in Fig. 5a. Confirming results from the dimensionality reduction analysis (Fig. 4), we found odor-evoked responses across different trials of an odorant at a particular intensity were highly correlated (diagonal blocks on each panel). The similarity between odor-evoked responses evoked by different concentrations of the same odorant was considerably less correlated (off-diagonal blocks). A hierarchical clustering analysis revealed that the odor-evoked responses were nicely clustered based on odor identity and then by intensity (Fig. 5b). These observations were further corroborated by a classification analysis (Supp. Fig. 4).

High and low stimulus intensities activate distinct ensembles.

(a) Correlations between neural responses observed in different trials are shown. Each pixel/matrix-element represents a similarity between mean neural responses in one trial versus those in another trial. Diagonal blocks reveal the correlation between trials when the same odorant at a specific intensity was repeatedly presented.

(b) A dendrogram was generated using a correlation distance metric comparing trial-by-trial ensemble spiking activities evoked by two different stimuli at two different intensities (see Methods). Two major response clusters that correspond to stimulus identity and intensity were identified. The number at the leaf node represents the trial number.

In sum, our results reveal that a combinatorial code could encode information regarding odor identity and intensity in an adaptation-invariant fashion. Since the same information about a stimulus is represented with fewer spikes in the later trials, we conclude that adaptation refines the odor codes by making them more efficient.

Neural response suppression vs. behavioral output facilitation

Neural responses evoked by an odorant reduced with repetition (Fig. 3). However, as we noted earlier, the behavioral responses increased upon stimulus repetition (Fig. 1). This negative correlation is evident when the neural responses (total spiking activities evoked across neurons) was plotted against the behavioral POR response probability during a particular trial (Fig. 6). This negative correlation between neural suppression and behavioral enhancement was observed for all odorants and both concentrations examined. It is worth noting that the last trial of the high-intensity exposure evoked neural responses that were comparable in strength to the first trial of the low-intensity exposure of the same stimulus. Yet, the behavioral responses observed during these two trials were markedly different. Hence, our results also indicate that the probability of the POR response is not simply a function of the total number of spikes elicited.

Neural adaptation inversely correlates with behavioral facilitation.

(a) The probability of odor-evoked POR (P(POR)) for a given trial is plotted against the total spike counts elicited during the trial or repetition number. Therefore, each point represents a single trial. Symbols with light and dark colors are used for differential trials of high and low-concentration odor exposures. The line represents the regression fit between the behavioral and neural responses (hex (L), R2 = 0.65; hex (H), R2 = 0.79). The negative slope of the regression line indicates that while the neural responses diminish over trials, the behavioral responses increase over trials.

(b-d) Similar plots as in panel a but for oct (L/H), iaa (L/H), and bza (L/H).

Neural and behavioral changes due to adaptation are odor-specific

Finally, we wondered if the neural response reduction and behavioral enhancements were a global, non-specific state change in the olfactory system brought about by the repetition of any odorant, or are the observed neural and behavioral response changes odor-specific. To examine this, we used a ‘catch-trial paradigm.’ (Fig. 7a) In these sets of experiments, we repeatedly presented an odorant to induce adaptation. After a substantial reduction in the neural response or stabilization of the behavioral response, a deviant or catch stimulus was presented. Persistence of the observed reduction in neural responses and enhancement in behavioral responses when the catch stimulus is introduced would reveal that the adaptation-induced changes are not odor-specific.

Adaptation results in odor-specific neural response reductions and behavioral output facilitation.

(a) Two blocks of trials were used. First, a block of 30 trials where one odorant (hex) was presented in all trials except the 26th trial (the catch trial). During the catch trial, a deviant stimulus (iaa or app) was presented. After a 15-minute no-odor reset period, a second block of ten trials of the deviant stimulus was presented (trials 31 – 40). This was done to determine the unadapted (first trial) and adapted (later trials) responses of the same set of PNs to the stimulus used in the catch trial.

(b) Summed spike counts (during 4 s odor presentation period) across all PNs were calculated and shown as a function of trials (hex repetitions). Repeated presentations of odor A resulted in a substantial reduction in odor-evoked spike counts. However, the presentation of the deviant odor (iaa) during the catch trial (trial # 26) resulted in a marked increase in spike counts. The response strength was similar to the non-adapted responses (Trial # 31) for iaa (H). (right) Similar results are shown when hex was repeatedly presented and the app was used as the deviant stimulus. Note that the response to the app during the catch trial did not recover as was noted for iaa (left).

(c) Correlations between odor-evoked responses with hex response in the first trial are shown as a function of trial number.

(d) A similar catch-trial paradigm was followed for behavioral experiments. Each locust was presented with a repeating odorant for eight trials. A deviant stimulus was presented in the ninth trial. This was followed by six more repetitions of the recurring stimulus.

(e) The response matrix summarizing the POR responses of individual locust PORs is shown. Same convention as in Fig. 1.

(f) The P(POR) across locusts was calculated and plotted as a function of the trial number. Notably, when the deviant stimulus was presented, there was a marked decrease in P(POR). The dotted line indicates the unadapted P(POR) value for the deviant stimulus.

We used the catch-trial experiments with two different catch stimuli. Our neural recordings reveal that when iaa (ester) was used as the catch stimulus after repeated presentations of hex (alcohol), the neural response strength recovered and increased to an overall higher spike count (Fig. 7b). The observed catch-iaa response strength matched the unadapted response level to iaa when it was separately presented following a 15-minute reset window (i.e., Fig. 7b, trial# 31). In contrast, when apple (a complex mixture) was used as the catch stimulus, the neural responses did not recover to higher levels indicating significant cross-adaptation (Fig. 7b). Nevertheless, in both cases, the ensemble responses evoked by the catch stimulus were highly similar to the unadapted responses it usually generates (Fig. 7c). Therefore, these results indicate that there is stimulus specificity to the adaptation-induced changes in neural responses and that cross-adaptation does not corrupt the identity of the deviant stimulus.

Behaviorally, we chose odor pairs where the component odorants had markedly different levels of POR responses (Fig. 7d). Repeated presentations of an odorant (bza or iaa) increased the p(POR) across the locusts as observed before (Fig. 7e). Notably, when the catch stimulus was presented, the behavioral responses dropped close to the first-trial p(POR) elicited by that odorant (Fig. 7f; dotted line). These findings were also observed when the recurring and catch odorants were switched (Fig. 7f).

In sum, these neural and behavioral results indicate that the observed neural and behavioral response changes brought about by the repeated encounters with an odorant are at least partially odor-specific and are not global, non-specific changes in the olfactory system.

Discussion

Sensory systems often have to deal with constraints that are of an opposing nature. For example, the ability to maintain stable representations is key to robust recognition of sensory stimuli42–44. Often, variations in the external environment, such as humidity and temperature, may introduce changes in neural responses that have to be filtered out to preserve the identity of the stimuli. Additionally, the strength or intensity of the stimulus can also change widely. Compressing or even removing these variations might be necessary to recognize an odorant independent of its intensity42–47. Alternately, in certain cases, it might also become important to maintain some of these variations so they can be exploited for guiding or altering the behavioral response. As an example, it might be worth noting that many chemicals that are pleasant at lower intensities can become repelling or even harmful to the organism at extremely high intensities13–15. Hence, the behavioral responses should accordingly vary with intensity. Further, many organisms have shown that they can follow an odorant to its source (source localization). Such behaviors are of ecological importance as they allow the organisms to find the food source or potential mates. In all these cases, it becomes important to maintain variations with stimulus intensity to ensure the organisms are able to approach the odor source. In our earlier studies, we have focused on examining how robust odor recognition is achieved37,40,48. Here, we examined the latter problem of how to maintain information about the stimulus intensity while the responses in the neural circuits are becoming weaker due to adaptation.

Consistent with earlier studies30,49,50, our data revealed that the response of individual projection neurons in the antennal lobe varied with the repetition of the same stimulus. While a reduction in odor-evoked responses was observed in many neurons, an increase in spiking activity and changes in temporal firing patterns were also observed. Notably, these changes were not random but highly repeatable (Fig. 2c). Hence, these results indicate that the effect of adaptation in a neural circuit is highly consistent and reliable.

In addition to spike rate reduction, oscillatory synchronization, and inter-neuronal coherence build up in the locust antennal lobe over repeated encounters with an odorant30,51,52. This short-term memory was not observed in the antenna but endured in the antennal lobe for several minutes after the termination of the stimulus. Similar results were also reported in the mouse olfactory bulb43. The inputs from sensory neurons were shown to be consistent across repetition, whereas the outputs of the olfactory bulb were systematically reduced with stimulus repetition. These results suggest that loci of short-term adaptation, at least when the stimulus is discontinuous and recurring, are in the antennal lobe/olfactory bulb circuitry. Although the precise mechanism for achieving the same is not fully understood49,53.

What is the computational significance of this neural adaptation? Sensory adaptation has been implicated in several important computations such as high-pass filtering, matching neural response to stimulus statistics, generating sparser codes, and optimal representations33,35,54–56. In the olfactory system, adaptation has also been suggested to stabilize neural representation such that recognition of odor identity from ensemble neural responses improved after repeated encounters with the stimulus30. However, adaptation also introduces a potential confound whereby the absolute neural response strength becomes an unreliable indicator of stimulus strength. Our results extend these earlier studies by revealing how this potential confound could be resolved in the olfactory system.

Increases in stimulus intensity have also been shown to result in higher response amplitudes in firing neurons and the recruitment of additional activated neurons35,42,44,57. While our results show that this is indeed true, we also found that several neurons responded preferentially during the lower concentration exposures but did not respond when the higher intensity of the same stimulus was presented (Supp. Fig. 5). Hence, the activated combination of neurons differed both with stimulus identity and intensity. Furthermore, even though adaptation reduced the overall spiking activities across neurons, the combination of neurons activated was robustly maintained. In other words, the population response vector direction (determined by the combination of neurons activated) was consistent. Only the vector length (determined by the total spikes across neurons) was reduced with the repetition of the stimulus. This simple approach was sufficient to encode the information about the odor intensity in an adaptation-invariant fashion. Additionally, since equivalent information about stimulus identity and intensity was preserved with fewer spikes, neural adaptation can be considered as a method of optimizing the neural representation2,3,9,10,33,34.

Does the observed persistence of concentration information in the neural activities after adaptation translate to observable differences in behavioral outcomes? Using an innate palp-opening response (POR), we examined how behavioral responses varied with stimulus repetition. Unlike the neural responses, we found that the PORs increased with repetition and, after about five trials, settled into odor-specific levels of responses. Notably, the PORs in the later trials were significantly higher for higher stimulus intensities for most of the odorants used. No such changes were observed during repeated presentation of paraffin oil (the solvent used for diluting odorants; Supp. Fig. 2). This suggests that the increases in behavioral responses with repetition were driven by the presence of the olfactory stimuli and not just the mechanical component (i.e. flow or pressure changes associated with stimulus presentations). Furthermore, our results showed that locusts are better able to behaviorally differentiate stimulus intensity after adaptation.

Alternately, adaptation due to the repetition of an odorant could alter the perceived intensity of the recurring odorant or other related test odorants. The compromise of stimulus intensity information due to adaptation has been observed in behavioral discrimination tasks in rats and even in psychophysical studies in humans58,59. Such a result can arise when the combinatorial response patterns evoked by the adapting and the test stimulus have significant overlap. However, our results indicate that when changes in odor intensities alter the combination of neurons activated, it is feasible to maintain information about odor intensity levels in a robust fashion.

Finally, we note that the anticorrelated link between neural suppression and behavioral enhancements (called ‘repetition priming’) is well established in other animal models60. How can weaker neural responses lead to enhanced behavioral outcomes? Our results indicate that the antennal lobe neural responses in later trials were not sharpened by pruning out weaker responses (Supp. Fig. 5). The weakening of the response strength could potentially lead to a sparser response in the downstream mushroom body. In addition, as we noted earlier, neural response synchrony increases with stimulus repetition. Whether weaker but synchronized activity in neurons drives sparser and more selective responses which then produce enhanced behavioral outcomes remains to be carefully examined.